A Guide to Designing Experiments to Test Statistical Graphics

### 2025-06-04 index.qmd
Add references and make some minor fixes to language. Convert refs.bib to ASCII.

Developing a graphics experiment is often a highly iterative process, but it can help to approach the design process by first optimizing the model and data generation method before spending time on optimizing the specific stimuli or customizing the data collection platform.
This is important because the model parameters and data generation process inform the experiment structure and thus impact decisions made downstream.

@@

One reason why it is so difficult to generate sampling models for visual explorations is that our visual system is optimized for identifying differences between groups.
This ability can interfere with the natural  to use the null sampling models that might be used in equivalent numerical tests when running experiments that use visualizations. 
We re-ran the experiment using a different clustering method that controlled the number of points in each group. 
Instead of noticing the number of ellipses, participants instead used the differences in size and shape of the ellipses formed when clustering after the data generating procedure. 
That is, participants could still detect the artificial nature of the induced clusters using other features. 
While it can be difficult to get the data generating method right, it is essential to conducting visual experiments that generalize well beyond the effects shown in a single data set or phenomenon.
This is also why it is critical to include independent replications of the simulated parameters, so that the results reflect variability due to the data generation process, not just the specifics of a single simulated dataset.
The time and effort invested in this step at the outset of the experiment pays dividends when it allows for clear generalization of the experimental results to an entire statistical concept rather than a single data set.

@@

>>>>

Developing a graphics experiment is often a highly iterative process, but it can help to approach the design process by first optimizing the model and data generation method (@sec-model-dev) before spending time on optimizing the specific stimuli or customizing the data collection platform (@sec-protocol-dev).
This is important because the model parameters and data generation process inform the experiment structure and thus impact decisions made downstream.

@@

One reason why it is so difficult to generate sampling models for visual explorations is that our visual system is very good at creating groups and identifying differences between them [@zekiFunctionalSpecializationGeneralization2013;@lupyanConceptualGroupingEffect2008].
This ability can interfere with the natural inclination to use the null sampling models that might be used in equivalent numerical tests when running experiments that use visualizations. 
Numerical tests consider one facet of the data; the power of graphical testing is that it allows examination of a battery of hypotheses simultaneously. 
Unfortunately, that makes designing an effective null sampling model much more difficult than in the numerical test case, as the researcher must effectively control for several different hypotheses when generating null data.

We re-ran the experiment using a different clustering method that controlled the number of points in each group. 
Instead of noticing the number of ellipses, participants instead used the differences in size and shape of the ellipses formed when clustering after the data generating procedure. 
That is, participants could still detect the artificial nature of the induced clusters using other features. 
While it can be difficult to get the data generating method right, it is essential to conducting visual experiments that generalize well beyond the effects shown in a single data set or phenomenon.
This is also why it is critical to include independent replications of the simulated parameters, so that the results reflect variability due to the data generation process, not just the specifics of a single simulated dataset.
The time and effort invested in this step at the outset of the experiment pays dividends when it allows for clear generalization of the experimental results to an entire statistical concept rather than a single data set.

@@

Once the pilot data has been collected and the planned analysis method has been developed, it may be beneficial to pre-register the main study and proposed analysis. 
This can be an effective way to separate exploratory research from confirmatory research involving hypothesis testing. 
However, graphics studies (and other human subjects research) often produce unexpected results that would not be imagined during the pre-registration process (for an example, see @clusters); using only the pre-registered analysis method may disincentivise analyzing (or collecting) the rich, messy open-ended response data which can be incredibly useful but is hard to model and quantify.
